Learning Auditory Similarities Between Instrumental Playing Techniques

### 2019-10-09 paper/paper.tex
write on the confusion matrices of patil

On the one hand, most publications in music information retrieval cast timbre modeling as an audio classification problem  \cite{martin1998asa,brown1999jasa,eronen2000icassp,herrera2003jnmr,wieczorkowska2003jiis,livshin2004dafx,krishna2004icassp,kaminskyj2005jiis,benetos2006icassp,bhalke2016jiis}.
In this context, the instrumentation of each musical excerpt serves as an unstructured set of ``audio tags'', encoded as binary outputs within some predefined label space.
Because such tags often belong to the publicly available metadata of music releases, the process of curating a training set for musical instrument classification requires little or no human intervention.
Although scraping user-generated content from online music hosting platforms may not always reflect the true instrumentation with perfect accuracy, it offers a scalable and ecologically valid insight onto the acoustic underpinnings of musical timbre.
Furthermore, supplementing user-generated content with the outcome of a crowdsourced annotation campaign allows an explicit verification of instrument tags.
For instance, the Open-MIC dataset \cite{humphrey2018ismir}, maintained by the Community for Open and Sustainable Music Information Research (COSMIR) \cite{mcfee2016ismir}, comprises a vast corpus of 20k polyphonic music recordings spanning across 20 instruments, as a derivative of the Free Music Archive (FMA) dataset \cite{defferrard2017ismir}.
Another example is the Medley-solos-DB dataset \cite{lostanlen2016ismir}, which comprises 21k monophonic excerpts from 8 instruments, as a derivative of the MedleyDB dataset of multitrack music \cite{bittner2014ismir}.
Over the past decade, the surge of these large digital audio collections, together with the democratization of high-performance computing on dedicated hardware, has spurred the development of deep learning techniques, and notably deep convolutional networks, in music instrument recognition \cite{mcfee2015ismir,pons2017eusipco,gururani2018ismir}.
We refer to \cite{han2017taslp} for a recent review of the state of the art in this domain.

On the other hand, the field of music psychology investigates timbre with the aim of discovering its physiological and behavioral foundations \cite{mcadams2009chapter}.
In this setting, prior knowledge, however accurate, of instrumentation does not suffice to conduct a study.
Rather, each excerpt must be played back to multiple human listeners.
Yet, collecting subjective responses to acoustic stimuli is a tedious and unscalable procedure, thus restricting the size of the musical corpus under study.
Such restriction hampers the applicability of iterative representation learning algorithms, such as stochastic gradient descent in artificial neural networks.
Nevertheless, advanced methods in electrophysiology allow to observe the firing patterns of biological neurons in the presence of controlled stimuli.
This observation, originally carried out on the ferret, has led to a comprehensive mapping of the primary auditory cortex, in terms of its spectrotemporal receptive fields (STRFs) \cite{depireux2001jneur}.
The STRF of a neuron is a function of time and frequency which represents the optimal predictor of its post-stimulus time histogram (PSTH) during exposure to a diverse range of auditory stimuli \cite{aertsen1981biolcyb}.
The simplest method to compute it in practice is by reverse correlation, i.e. by averaging all stimuli that trigger an action potential \cite{deboer1968biomed}.
Historically, STRFs were defined by their Wigner--Ville distribution \cite{flandrin1998book}, thereby sparing the choice of a tradeoff in time--frequency localization, but suffering from cross-term interference \cite{eggermont1993hearing}.
Since then, the STRF of a neuron was redefined as a spectrographic representation of its spike-triggered average \cite{klein2000compneur}.
Although this new definition is necessarily tied to a choice of spectrogram parameters, it yields more interpretable patterns than a Wigner-Ville distribution.
In particular, a substantial portion of spectrographic STRFs exhibits a ripple-like response around a given region $(t, \lambda)$ of the time--frequency domain.
This response can be approximately described by a pair of scalar values: a temporal modulation rate $\alpha$ in Hertz and a frequential modulation rate $\beta$ in cycles per octave.
Interestingly, both $\alpha$ and $\beta$ appear to be arranged in a geometric series and independent from the selected center time $t$ and center frequency $\lambda$.
This observation has led auditory neuroscientists to formulate an idealized computational model for STRF, known as ``full cortical model'' \cite{chi2005jasa}. 
Over recent years, several publications have employed the full cortical model as a feature extractor for a task of musical instrument classification, both in isolated recordings \cite{patil2012ploscompbiol} and in solo phrases \cite{patil2015eurasip}.
These biologically inspired features have shown to outperform the state of the art, especially in the small data regime.

>>>>

On the one hand, most publications in music information retrieval cast timbre modeling as an audio classification problem  \cite{martin1998asa,brown1999jasa,eronen2000icassp,herrera2003jnmr,wieczorkowska2003jiis,livshin2004dafx,krishna2004icassp,kaminskyj2005jiis,benetos2006icassp,bhalke2016jiis}.
In this context, the instrumentation of each musical excerpt serves as an unstructured set of ``audio tags'', encoded as binary outputs within some predefined label space.
Because such tags often belong to the metadata of music releases, the process of curating a training set for musical instrument classification requires little or no human intervention.
Although scraping user-generated content from online music hosting platforms may not always reflect the true instrumentation with perfect accuracy, it offers a scalable and ecologically valid insight onto the acoustic underpinnings of musical timbre.
Furthermore, supplementing user-generated content with the outcome of a crowdsourced annotation campaign allows an explicit verification of instrument tags.
For instance, the Open-MIC dataset \cite{humphrey2018ismir}, maintained by the Community for Open and Sustainable Music Information Research (COSMIR) \cite{mcfee2016ismir}, comprises a vast corpus of 20k polyphonic music recordings spanning across 20 instruments, as a derivative of the Free Music Archive (FMA) dataset \cite{defferrard2017ismir}.
Another example is the Medley-solos-DB dataset \cite{lostanlen2016ismir}, which comprises 21k monophonic excerpts from 8 instruments, as a derivative of the MedleyDB dataset of multitrack music \cite{bittner2014ismir}.
Over the past decade, the availability of large digital audio collections, together with the democratization of high-performance computing on dedicated hardware, has spurred the development of deep learning architectures, and notably deep convolutional networks, in music instrument recognition \cite{mcfee2015ismir,pons2017eusipco,gururani2018ismir}.
Notwithstanding the growing accuracy of these architectures in the large data regime, it remains unclear how to extend them from musical instrument recognition to playing technique recognition, where labeled samples are considerably scarcer.
We refer to \cite{han2017taslp} for a recent review of the state of the art in this domain.

On the other hand, the field of music psychology investigates timbre with the aim of discovering its physiological and behavioral foundations \cite{mcadams2009chapter}.
In this setting, prior knowledge, however accurate, of instrumentation does not suffice to conduct a study.
Rather, each excerpt must be played back to multiple human listeners.
Yet, collecting subjective responses to acoustic stimuli is a tedious and unscalable procedure, thus restricting the size of the musical corpus under study.
Such restriction hampers the applicability of iterative representation learning algorithms, such as stochastic gradient descent in artificial neural networks.
Nevertheless, advanced methods in electrophysiology allow to observe the firing patterns of biological neurons in the presence of controlled stimuli.
This observation, originally carried out on the ferret, has led to a comprehensive mapping of the primary auditory cortex, in terms of its spectrotemporal receptive fields (STRFs) \cite{depireux2001jneur}.
The STRF of a neuron is a function of time and frequency which represents the optimal predictor of its post-stimulus time histogram (PSTH) during exposure to a diverse range of auditory stimuli \cite{aertsen1981biolcyb}.
The simplest method to compute it in practice is by reverse correlation, i.e. by averaging all stimuli that trigger an action potential \cite{deboer1968biomed}.
Historically, STRFs were defined by their Wigner--Ville distribution \cite{flandrin1998book}, thereby sparing the choice of a tradeoff in time--frequency localization, but suffering from cross-term interference \cite{eggermont1993hearing}.
Since then, the STRF of a neuron was redefined as a spectrographic representation of its spike-triggered average \cite{klein2000compneur}.
Although this new definition is necessarily tied to a choice of spectrogram parameters, it yields more interpretable patterns than a Wigner-Ville distribution.
In particular, a substantial portion of spectrographic STRFs exhibits a ripple-like response around a given region $(t, \lambda)$ of the time--frequency domain.
This response can be approximately described by a pair of scalar values: a temporal modulation rate $\alpha$ in Hertz and a frequential modulation rate $\beta$ in cycles per octave.
Interestingly, both $\alpha$ and $\beta$ appear to be arranged in a geometric series and independent from the selected center time $t$ and center frequency $\lambda$.
This observation has led auditory neuroscientists to formulate an idealized computational model for STRF, known as ``full cortical model'' \cite{chi2005jasa}. 
Over recent years, several publications have employed the full cortical model as a feature extractor for a task of musical instrument classification, both in isolated recordings \cite{patil2012ploscompbiol} and in solo phrases \cite{patil2015eurasip}.
These biologically inspired features have shown to outperform the state of the art, especially in the small data regime when deep learning is inapplicable. % TODO cite?
Furthermore, the confusion matrix of the full cortical model in the label space of musical instruments strongly correlates with the confusion matrix between a human listener and the ground truth.
In other words, although there is no evidence that the full cortical model consistently predicts the same instrument labels as a human listener, cross-instrument confusions do occur in the same proportions overall.
